Computing in the Field of Rationals

### 2012-06-21 rational_ext.tex
_

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \fatsemi
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftP = L^+
%! RightP = R^+
%! LeftT = L^{\times}
%! RightT = R^{\times}
%! alpha = \alpha
%! bool = \textit{bool}
%! color = \textit{color}
%! Gr = G

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

\begin{lemma}[{{Pi}} is a category]
  We have as objects the types {{b}} and as morphisms equivalence
  classes of combinators {{b1<->b2}}. One can check:
  \begin{enumerate}
  \item Every object {{b}} has an identity morphism {{b <->b}}.
  \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
    {{g:b2<->b3}} is given by sequencing {{f (;) g}}. 
  \item Associativity of composition follows from operational
    equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
    {{h:b3<->b4}}). One can check:

@@

\begin{lemma}[Dagger]
  {{Pi}} is a dagger category, where every morphism {{f : b1<->b2}}
  has the adjoint {{f^{dagger}:b2 <->b1}}. The following properties
  hold (where {{g : b2 <-> b3}}): 
  \begin{enumerate}
  \item {{id^{dagger} = id : b <-> b}}
  \item {{(f (;) g)^{dagger} = f^{dagger} (;) g^{dagger}: b1 <-> b2}}
  \item {{f^{dagger dagger} = f : b1 <-> b2}}
  \end{enumerate}
\end{lemma}

@@

\begin{lemma}[Symmetric Monoidal (*, 1)]
  {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
  unit {{1}}. The details here follow those of the {{(0, +)}} monoid
  and are skipped.
\end{lemma}

Some technical and pedantic comments are due at this point.

@@

\item The category {{Pi}} has no initial and terminal objects. The
  objects {{0}} and {{1}} would be initial and terminal if we admitted
  all functions (is in the category {{CatSet}}).

\item The category {{Pi}} has neither categorical products, nor
  categorical co-products -- {{*}} and {{+}} are merely monoidal
  tensors. This follows from the fact that injection and projection
  maps are not isomorphisms.
\end{itemize}

While we don't do so in this work, if we extend {{Pi}} with products
and co-products (say through the addition of information effects) it
is concievable that part of its structure collapses. The sort of
collapse, while catastrophic for algebraic structures (effectively
trivializing them) still retains some interest in computing, because in
computing we are interested in the specific operational nature (the
computation content, so to speak) of the morphisms. Anectodal evidence
follows from the fact several real-world programming language have
inconsistent type systems. blah blah blah...

>>>>

%subcode-inline{bnf-inline} name langRev
%! swap+ = \mathit{swap}^+
%! swap* = \mathit{swap}^*
%! dagger =  ^{\dagger}
%! assocl+ = \mathit{assocl}^+
%! assocr+ = \mathit{assocr}^+
%! assocl* = \mathit{assocl}^*
%! assocr* = \mathit{assocr}^*
%! identr* = \mathit{uniti}
%! identl* = \mathit{unite}
%! dist = \mathit{distrib}
%! factor = \mathit{factor}
%! eta = \eta
%! eps = \epsilon
%! eta+ = \eta^+
%! eps+ = \epsilon^+
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! trace+ = trace^+
%! trace* = trace^{\times}
%! ^^^ = ^{-1}
%! (o) = \circ
%! (;) = \fatsemi
%! (*) = \times
%! (+) = +
%! LeftP = L^+
%! RightP = R^+
%! LeftT = L^{\times}
%! RightT = R^{\times}
%! alpha = \alpha
%! bool = \textit{bool}
%! color = \textit{color}
%! Gr = G

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! CatSet = \textbf{Set}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

We present the categorical structure with minimum commentry. More
details may be found in excellent references such as Barr and Wells
[CITE] and Selinger[CITE].

\begin{lemma}[{{Pi}} is a category]
  The category {{Pi}} has the types {{b}} as objects and equivalence
  classes of well-typed combinators {{b1<->b2}} as morphisms. One can
  check:
  \begin{enumerate}
  \item Every object {{b}} has an identity morphism {{id : b <->b}}.
  \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
    {{g:b2<->b3}} is given by sequencing {{f (;) g}}. 
  \item Associativity of composition follows from operational
    equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
    {{h:b3<->b4}}). One can check:

@@

\begin{lemma}[Dagger]
  {{Pi}} is a dagger category, where every morphism {{f : b1<->b2}}
  has the adjoint {{f^{dagger}:b2 <->b1}}. The following properties
  hold (where {{g : b2 <-> b3}}): 
  \begin{enumerate}
  \item {{id^{dagger} = id : b <-> b}}
  \item {{(f (;) g)^{dagger} = g^{dagger} (;) f^{dagger}: b3 <-> b1}}
  \item {{f^{dagger dagger} = f : b1 <-> b2}}
  \end{enumerate}
\end{lemma}

@@

\begin{lemma}[Symmetric Monoidal {{(*, 1)}}]
  {{Pi}} is a symmetric monoidal category over the tensor {{*}} and
  unit {{1}}. The details mirror those of the {{(0, +)}} monoid.
\end{lemma}

\noindent
Some technical and pedantic comments are due at this point.

@@

\item The category {{Pi}} has no initial and terminal objects. The
  objects {{0}} and {{1}} would be initial and terminal if we admitted
  all functions (as in the category {{CatSet}}).

\item The category {{Pi}} has neither categorical products, nor
  categorical co-products, i.e. {{*}} and {{+}} are merely monoidal
  tensors. This follows from the fact that injection and projection
  maps are not isomorphisms.
\end{itemize}

While we don't do so in this work, if we extend {{Pi}} with products
and co-products (say through the addition of information effects) it
is concievable that part of its structure collapses. This sort of
collapse, while catastrophic for algebraic structures (effectively
trivializing them), still retains some interest in computing, because
in computing we are interested in the specific operational nature (the
computational content, so to speak) of the morphisms. Anectodal evidence
follows from the fact several real-world programming language have
inconsistent type systems. blah blah blah...
